---
title: "ESPM 154, Lab 2: Neutral Landscape Models"
output:
  html_document:
    toc: true
    toc_float: true
---

## Overview and learning goals

Landscape patterns can arise from many biological and environmental processes. Neutral landscape models (NLMs) give us landscapes with known statistical properties but no assumed ecological process. They provide a useful baseline for asking whether a pattern is unusual and for testing how spatial structure affects an analysis.

In this lab, you will:

1. Generate simple random and spatially autocorrelated neutral landscapes.
2. Interpret semivariograms and Moran's I.
3. Inspect the spatial scale of a real environmental raster.
4. Test how changing raster grain changes a spatial statistic.

## Set up the R session

```{r setup, message=FALSE, warning=FALSE}
rm(list = ls())

if (!requireNamespace("terra", quietly = TRUE)) {
  stop("The terra package is required. Run install.packages('terra') in the Console, then restart this document.")
}

if (!requireNamespace("ambient", quietly = TRUE)) {
  stop("The ambient package is required. Run install.packages('ambient') in the Console, then restart this document.")
}

library(terra)
library(ambient)
knitr::opts_chunk$set(echo = TRUE, fig.width = 8, fig.height = 5)

source("NLM_functions.R")

nlm_colors <- hcl.colors(100, "Viridis")
temp_colors <- hcl.colors(100, "Inferno")


# Queen neighborhood: the eight cells surrounding a focal cell.
queen_weights <- matrix(c(1, 1, 1,
                          1, 0, 1,
                          1, 1, 1), nrow = 3, byrow = TRUE)
```

## Part 1: Neutral landscape models

### Simple random landscape

The simplest NLM assigns each cell an independent random value. It has no built-in spatial structure, although a finite random landscape will rarely have Moran's I exactly equal to zero.

Generate and plot a simple random NLM with 200 rows and 200 columns.

```{r simple-nlm}
simple <- nlm_simple(ncol = 200, nrow = 200)
plot(simple, col = nlm_colors, axes = FALSE, main = "Simple random NLM")
```

### Spatially autocorrelated landscapes

The `nlm_fractal()` function combines random variation generated at several spatial scales. Here, `H` controls the relative importance of broad-scale structure. Larger values produce smoother landscapes with stronger positive spatial autocorrelation.

Generate a landscape with `H = 0.9`.

```{r fractal-high, warning = F}
# Generate a fractal NLM
fractal_0.9 <- nlm_fractal(ncol = 200, nrow = 200, H = 0.9)

# Plot the NLM
plot(fractal_0.9, col = nlm_colors, axes = FALSE, main = "H = 0.9")
```

Now complete the code to generate a landscape with `H = 0.3`. Use the same seed so that the comparison emphasizes the effect of `H`.

```{r fractal-low, warning = F}
fractal_0.3 <- ### Complete this line
```

Plot the two landscapes using a common color scale.

```{r compare-fractals, fig.width=9, fig.height=4}
fractals <- c(fractal_0.3, fractal_0.9)
plot(fractals, col = nlm_colors, axes = FALSE,
     range = c(0, 1), nc = 2)
```

### Semivariograms

A semivariogram summarizes how dissimilarity changes with distance. Low semivariance at short distances indicates that nearby cells tend to have closer values. A rapid increase indicates that variation between cells increases quickly over relatively short distances. If the curve reaches a plateau (the `sill`), the approximate distance where it levels off is the `range distance`.

```{r semivariograms, fig.width=9, fig.height=4}
var_0.3 <- semivariogram(fractal_0.3)
var_0.9 <- semivariogram(fractal_0.9)

par(mfrow = c(1, 2))
plot_variogram(var_0.3, main = "Semivariogram H = 0.3")
plot_variogram(var_0.9, main = "Semivariogram H = 0.9")
par(mfrow = c(1, 1))
```

Use these plots to answer *Question 1.* 

### Moran's I

We can calculate Moran's I using the `autocor()` function. When calculating Moran's I, we have to specify which cells are considered neighbors. Here we will use a `queen's neighborhood` (all eight directions from each cells). 

```{r moran-simple}
simple_I <- autocor(simple, w = queen_weights, method = "moran", global = TRUE)
simple_I
```

Calculate Moran's I for the landscape with `H = 0.9`.

```{r moran-high}
H09_I <- autocor(fractal_0.9, w = queen_weights, method = "moran", global = TRUE)
H09_I
```

Now complete the code to calculate Moran's I for the landscape with `H = 0.3`.

```{r moran-low}
H03_I <- ### Complete this code
H03_I
```

Use these results to answer *Question 2.*


## Part 2: A real-world temperature raster

Load annual mean temperature (AMT) for California and part of Nevada. `rast()` creates a `SpatRaster`, the raster data class used by `terra`.

```{r load-temperature}
# Load the raster
AMT <- rast("CA_AMT.tif")

# Look at what's contained in the SpatRaster object.
AMT
```

```{r plot-temperature}
plot(AMT, col = temp_colors, main = "Annual mean temperature")
```

Calculate Moran's I for this AMT raster using the same queen's neighborhood as above.

```{r moran-temperature}
AMT_I <- ### Complete this code
```

Use these results to answer *Question 3.*

## Part 3: Effects of changing scale

We can create coarser versions of the temperature raster by combining neighboring cells. Using the `aggregate()` function with the argument `fun = "mean"`, each new cell will contain the mean temperature of the cells it replaces.

```{r aggregate-temperature-4x}
# Aggregate cells by a factor of 4 (`fact = 4`)
AMT_4 <- aggregate(AMT, fact = 4, fun = "mean", na.rm = TRUE)
```

Now, rescale the AMT raster by aggregating its cells by a factor of 16 and by a factor of 64.

```{r}
AMT_16 <- ### Complete this code
AMT_64 <- ### Complete this code
```

Then we can plot the three AMT rasters.

```{r plot-AMTs}
# Plot all three AMT rasters
temperature_range <- range(values(AMT), na.rm = TRUE)
par(mfrow = c(1, 3))

plot(AMT,
     col = temp_colors,
     range = temperature_range,
     main = "Original")

plot(AMT_4,
     col = temp_colors,
     range = temperature_range,
     main = "4 x coarser")

plot(AMT_16,
     col = temp_colors,
     range = temperature_range,
     main = "16 x coarser")

par(mfrow = c(1, 1))
```


Now calculate Moran's I for all three rasters.

```{r compare-moran}
moran_by_grain <- data.frame(
  raster = c("Original", "4x coarser", "16x coarser", "64x coarser"),
  morans_I = c(
    AMT_I,
    autocor(AMT_4, w = queen_weights, method = "moran", global = TRUE),
    autocor(AMT_16, w = queen_weights, method = "moran", global = TRUE),
    autocor(AMT_64, w = queen_weights, method = "moran", global = TRUE)
  )
)

moran_by_grain
```

Use these results to answer Question 4. Remember that the same 3 by 3 neighborhood covers a larger geographic area after aggregation.

## End of lab
